---
title: "Introduction to BayGMST"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BayGMST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
has_cmdstan <- requireNamespace("instantiate", quietly = TRUE) &&
  instantiate::stan_cmdstan_exists()
```

This vignette walks through the full BayGMST workflow end to end, using small
example datasets bundled with the package (trimmed subsets of the data used
in Bagwell et al., in prep.; see `system.file("extdata", package =
"BayGMST")`). It replaces the `config.yml`-driven scripts the package was
originally built from -- see `NEWS.md` and
`system.file("legacy-scripts", package = "BayGMST")`.

```{r setup}
library(BayGMST)
```

## 1. Load example data

```{r load-data}
proxy_matrix <- read.csv(
  system.file("extdata", "example_proxy_matrix.csv", package = "BayGMST")
)
forcing_raw <- read.csv(
  system.file("extdata", "example_forcing.csv", package = "BayGMST")
)
instrumental_raw <- read.csv(
  system.file("extdata", "example_instrumental_temperature.csv", package = "BayGMST")
)
# read.csv() mangles "Time"/"Anomaly (deg C)" headers via make.names(); give
# them stable names instead of relying on the mangled spelling.
colnames(instrumental_raw)[1:2] <- c("Time", "Anomaly")

years <- forcing_raw$year
head(proxy_matrix)
```

## 2. Reduce the proxy network to a single representative series

```{r reduce-proxies}
calib_years <- instrumental_raw$Time

rp <- reduce_proxies(
  proxy_matrix = proxy_matrix[, -1],
  years        = proxy_matrix$year,
  temp_calib   = instrumental_raw$Anomaly,
  calib_years  = calib_years,
  method       = "PCR",
  # `chunk` must be large enough that every segment (including the last,
  # shortest one) still fully contains the calibration window (1850-2000,
  # 151 years). This toy example only has 301 years total, so we use a
  # single segment (chunk >= length(years)) rather than the package
  # default of 250, which would create a too-short trailing segment here.
  chunk        = nrow(proxy_matrix)
)
head(rp$composite)
```

`rp$composite$RP1` is the representative proxy series to pass to
`fit_baygmst()`.

## 3. Transform the radiative forcings

```{r transform-forcings}
forcing <- transform_forcings(
  G = forcing_raw$CO2,
  V = forcing_raw$volcanic,
  S = forcing_raw$solar
)
```

## 4. Align everything onto the common year grid and fit the model

```{r align}
instrumental_T <- instrumental_raw$Anomaly[match(years, instrumental_raw$Time)]
proxy <- rp$composite$RP1[match(years, rp$composite$year)]
```

Fitting requires a working CmdStan installation (via \pkg{cmdstanr} /
\pkg{instantiate}); see `vignette("cmdstanr", package = "cmdstanr")` if you
don't have one yet. The chunk below is skipped if none is found.

```{r fit, eval = has_cmdstan}
fit <- fit_baygmst(
  proxy          = proxy,
  instrumental_T = instrumental_T,
  forcing_G      = forcing$G,
  forcing_V      = forcing$V,
  forcing_S      = forcing$S,
  years          = years,
  chains         = 2,
  parallel_chains = 2,
  iter_warmup    = 300,
  iter_sampling  = 1000,
  seed           = 1
)
summary(fit)
```

## 5. Visualize the reconstruction

```{r plot-reconstruction, eval = has_cmdstan}
plot_reconstruction(fit)
```

```{r plot-trace, eval = has_cmdstan}
plot_trace(fit)
```

```{r plot-posteriors, eval = has_cmdstan}
plot_posterior_densities(fit)
```

## 6. Cross-validate

```{r cv, eval = has_cmdstan}
cv <- cv_baygmst(
  proxy          = proxy,
  instrumental_T = instrumental_T,
  forcing_G      = forcing$G,
  forcing_V      = forcing$V,
  forcing_S      = forcing$S,
  years          = years,
  nfold          = 3,
  chains         = 2,
  parallel_chains = 2,
  iter_warmup    = 300,
  iter_sampling  = 1000,
  seed           = 1
)
cv
plot_cv(cv)
```

## Known limitations

See `NEWS.md` for the full list. Most notably: `src/stan/baygmst.stan`
contains one unresolved indexing question the original author flagged and
never fully validated (`// CHECK THIS LINE`), carried forward unchanged
pending statistical review -- reconstructions from this vignette's toy data
should not be over-interpreted.
